Predicting Water Treatment Plant Chemical Doses

### 2025-12-10 Presentation DSCI445.Rmd
Made lots of changes to presentation.

---
title: "DSCI 445 Presentation"
author: "Ethan Schilling, Joshua Tobey, Dawson Carney, Reece Carmody"
date: "2025-12-11"
output: 
  beamer_presentation:
    theme: "Warsaw"
    colortheme: "default"
---

@@

```{r, include=F}
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(zoo)
library(purrr)
library(knitr)
library(patchwork)
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(kableExtra)
library(vip)
```

# Example Slide Title

Example text.

- Bullet points
- Bullet points

## New section of text

More example text

```{r}
#readRDS("TOC_plot.rds")
#readRDS("TDS_plot.rds")
```

@@

The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards. There are four primary steps of the water treatment process:

1. Coagulation - Chemicals ("coagulants") are added to raw water to help contaminants
group together into "floc" particles  
2. Flocculation - The water is then slowly mixed to allow these floc particles to grow  
3. Filtration - These particles are filtered out  
4. Disinfection - Water is disinfected to get rid of biological contaminants

This project will focus on Step 1. These chemical doses are one of the most important
features of the treatment process that an operator can change to improve performance.

@@

**The role of an operator in deciding water treatment plant chemical doses is to
look at a wide array of factors and make decisions about how to adjust
chemical doses.**

- Treatment plant conditions change rapidly
- A predictive tool to help select chemical doses is beneficial both financially and environmentally

```{r}
# Add explanation picture/plot to help people visualize here??
```

@@

Three years of data from 2018-2020, provided by a water treatment plant in Aurora, Colorado.

@@

- Features were averaged to match the dosing data's daily increments 
- Merged various datasets into a final cleaned dataset
- Various outliers (likely caused by equipment malfunction) were either removed or interpolated
- Suspended Grain Size was not used as it was highly variable
- Alkalinity and Hardness was not used because of missing data chunks and differing time ranges

@@

Generally, temperature and pH seem to have the strongest relationship. This coincides with feedback from treatment plant operators. For further analysis, we will use:

- pH
- Temperature
- Conductivity
- Turbidity

# Introducing Lagged Variables

Similar to the Stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.

@@

@@

**Add slides for various numerical prediction models**

@@

@@

# Coagulant Change Prediction

**Add slides for various classification models**

>>>>

---
title: "Predicting Water Treatment Plant Chemical Doses"
author: "Ethan Schilling, Joshua Tobey, Dawson Carney, Reece Carmody"
date: "11 December 2025"
aspectratio: 169
output: 
  beamer_presentation:
    theme: "Warsaw"
    colortheme: "default"
---

@@

```{r, include=F}
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(zoo)
library(purrr)
library(knitr)
library(patchwork)
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(kableExtra)
library(vip)
library(png)
library(grid)
```

@@

The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.

```{r}
img_array = readPNG("Images/treatmentplot1.png")
grid.raster(img_array)
```

@@

**One role of an operator in the water treatment process is to
look at a wide array of data and make decisions about how to adjust
chemical doses.**

```{r}
img_array = readPNG("Images/treatmentplot2.png")
grid.raster(img_array)
```

**OUR GOAL: Create a model that serves the role of an "operator," taking raw water data and making predictions of chemical doses.**

```{r}
img_array = readPNG("Images/treatmentplot3.png")
grid.raster(img_array)
```

@@

Three years of data from 2018-2020, provided by a Colorado water treatment plant.

@@

- Daily averages were taken for raw water data to match the dosing data's daily timestep 
- Merged various datasets into a final cleaned dataset
- Various large outliers (likely caused by equipment malfunction) were removed
- Temperature outliers were interpolated to nearest adjacent values
- Suspended grain size was not used as it was highly variable
- Alkalinity and hardness were not used because of missing data chunks and differing time ranges

@@

Temperature and pH seem to have the strongest relationship to dose, consistent with feedback from treatment plant operators. Other data were either highly variable or had minimum availability with large gaps. We will focus on the following to predict dose:

- pH
- Temperature
- Conductivity (lots of noise, weaker)
- Turbidity (lots of noise, weaker)

Focus: **FeCl Dose**. Polymer dose is not changed significantly in plants.

# Added Variables

Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.

@@

We also introduced a categorical predictor indicating whether the dose increased, decreased, or stayed the same compared to the day before.

# Summary of Models

**GOAL: Interpretable and Captures Nonlinear Behavior**

- Training and Testing
   - Training: 2018-2019
   - Testing: 2020
   - Approximately 70/30 split
- Coagulant Dose Prediction
   - Linear Regression
   - Tree-based Models
   - GAM
- Coagulant Dose Change Prediction
   - Multinomial Classification
   - LDA, QDA
   - Random Forest

@@

**GOAL: Predict coagulant doses based on raw water characteristics.**

```{r, fig.width=7, fig.height=3}

df_lag_incdec_long <- df_lag_incdec |>
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "Characteristics",
    values_to = "Value"
  )

ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value, color = FeCl_dose)) +
  geom_point(size = 0.7) +
  facet_wrap(~Characteristics, scales = "free_y") +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Value",
    color = "Ferric Chloride (mg/L)",
    title = "Raw Water Characteristics Over Time Colored by Dosing"
  )

# Coagulant Dose Prediction: Linear Regression

**Ethan: insert table with models attempted**

@@

```{r, fig.width=6.5, fig.height=3.4}
set.seed(445)

@@

# Coagulant Dose Change Prediction

**GOAL: Predict whether coagulant dose increased, decreased, or stayed the same based on raw water characteristics.**

```{r, fig.width=7, fig.height=3}

ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value, color = FeCl_change)) +
  geom_point(size = 0.7) +
  facet_wrap(~Characteristics, scales = "free_y") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Value",
    color = "FeCl_change",
    shape = "FeCl_change",
    title = "Raw Water Characteristics Colored by Dose Change"
  )

# Coagulant Dose Change Prediction: Modeling

#

# Summary of Dose Prediction Models

**Insert bar graph/brief conclusions**

# Summary of Dose Change Prediction Models

**Insert bar graph/brief conclusions**
